---
title: "Class 11 Introduction to Genome Informatics Lab"
author: "Julia Ainsworth"
format: html
editor: visual
---


#Section 1 Proportion of G/G in a population


Downloaded a CSV file from Ensembl < https://www.ensembl.org/Homo_sapiens/Variation/Summary?v=rs8067378;toggle_HGVS_names=open


Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```




```{r}
table(mxl$Genotype..forward.strand.)
```
Here is the number, but we want the percentage

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) * 100
```


Now, looking at the proportion in the British population GBR: 



```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (1).csv")
head(gbr)
```
Find proportion of G/G
```{r}
table(gbr$Genotype..forward.strand.)/nrow(gbr) *100
```

This variant is associated with childhood asthma and is found more frequently in gbr populations than mxl.

Let's dig into this further.







