---
title: "Class 18 Pertussis Mini Project"
format: pdf
editor: visual
toc: true
---

# Is Pertussis on the rise

Install packages - datapasta

don't need to Load library! We can click Addins button instead and paste as data frame - one way to scrape data from the internet. Downside is, we need to make sure to copy/paste accurately

Data copied from: https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html
```{r}
#not adding data frame to PDF
 cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617)
       )
```

# Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.
```{r}
library(ggplot2)
```


```{r}
baseplot <- ggplot(data = cdc) + 
  aes(x = Year, y = Cases) + 
  geom_line() + 
  labs(title = "Pertussis Cases Over Time in the US") 
  
baseplot
```

# Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
#Adding a vline for the introduction of the first pertussis vaccine in 1996 and the change in formulation of the vaccine in 1996

baseplot +
  geom_vline(xintercept = 1946, color = "red", linetype = 2) + geom_vline(xintercept = 1996, color = "blue", linetype =2)

```

# Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?
The pertussis strain could be mutating and the vaccine formulation could be less effective; alternately, the new vaccine formulation could be less effective versus the one formulated from the whole bacteria.  Or, it could be a decrease in vaccination rates due to the paper from 2006.  Evidence suggests that it's not variation in the bacteria that causes pertussis, but it might be waning efficacy due to the new formulation of the pertussis vaccine.

Installing the jsonlite package 
Loading in the library
```{r}
library(jsonlite)
```



Reading a json file from the website:
```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

# Q4. How many aP and wP infancy vaccinated subjects are in the dataset?
wP: 49, aP = 47
```{r}
table(subject$infancy_vac)
```

# Q5. How many Male and Female subjects/patients are in the dataset?
F: 66
M: 30

```{r}
table(subject$biological_sex)
```

# Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

# Dates 

```{r}
head(subject$year_of_birth, 3)
head(subject$date_of_boost, 3)
```

```{r}
library(lubridate)
today()
```

```{r}
today() - ymd("2000-01-01")
time_length( today() - ymd("2000-01-01"), "years")

```

Making a new column for age of all subjects in the trial
```{r}
subject$age <- time_length( today() - ymd(subject$year_of_birth), "years")
head(subject, 3)
```

# Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?
```{r}
summary(subject$age)
```
# Q8: 

# Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Making a histogram
```{r}
base_hist <- ggplot(subject) + 
  aes(age) + 
  geom_histogram()
base_hist
```
Facet wrapping by type of vaccine shows the point at which the aP vaccine was introduced 
```{r}
base_hist + 
  facet_wrap(vars(infancy_vac), nrow = 2)
```
Using dplyr to join different tables together - get specimen table and join with subject to have all information in one place

```{r}
library(dplyr)
```
Read the specimen table
```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 

```

look at specimen table: 
```{r}
head(specimen, 3)
```
# Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

Using a join function to bring the tables together. It will automatically find what's in common between the tables, which here is "subject_id", and joins them together -- same number of rows as in specimens, but with extra columns from subject
```{r}
library(dplyr)
```

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
```

Now reading some experimental data
```{r}
titer <- read_json("https://www.cmi-pb.org/api/ab_titer", simplifyVector = TRUE) 
head(titer)

```

# Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Join the titer table with the meta table! 
```{r}
abdata <- inner_join(meta, titer)
dim(abdata)

```

# Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```


# Q12. What do you notice about the number of visit 8 specimens compared to other visits?
Much fewer of them! So, we will drop this from our data, as well as focus in on one isotype - I will use IgG1
```{r}
table(abdata$visit)
```

Making a plot for titer levels of different antiboides 

```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot() 
```

Will define ig1 as dataset for only IgG1 and without visit 8:
```{r}
ig1 <- abdata %>%
  filter(isotype == "IgG1", visit != 8)
head(ig1)
  
```


# Q13. Complete the following code to make a summary boxplot of Ab titer levels for all antigens:
```{r}
ggplot(ig1) + 
  aes(MFI, antigen, col = infancy_vac) + 
  geom_boxplot(show.legend = FALSE) +
   facet_wrap(vars(visit), nrow=2) +
  theme_bw()

```

# Q14. What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?
TT, PTM, PT, PRN, FIM2/3, LOS, FHA, FELD1 and DT

These are all antigens in the vaccine, whereas antibodies to measles don't change because people in the study haven't been exposed to measles 




