---
title: "Optional Vaccination Rate Project"
author: "Julia Ainsworth"
format: pdf
toc: true
editor: visual
---

# Beginning: 

Importing vaccination data for San Diego:

```{r}
vax <- read.csv(file = "covid19vaccinesbyzipcode_test.csv")
# head(vax) note: was cluttering up pdf 
```

# Q1. What column details the total number of people fully vaccinated?

A: persons_fully_vaccinated

# Q2. What column details the Zip code tabulation area?

A: zip_code_tabulation_area

```{r}
head(vax$as_of_date)
tail(vax$as_of_date)
```

# Q3. What is the earliest date in this dataset?

A: 2021-01-05

# Q4. What is the latest date in this dataset?

A: 2022-11-22

```{r}
# loaded skimr library in the console
skimr::skim(vax)
```

# Q5. How many numeric columns are in this dataset?

A: 13

# Q6. Note that there are "missing values" in the dataset. How many NA values there in the persons_fully_vaccinated column?
14921
(code below)

# Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?
.00069 %

```{r}
a <- sum( is.na(vax$persons_fully_vaccinated) ) #Question 6
a
b <- sum(vax$persons_fully_vaccinated, na.rm = T)

(a/b) * 100 # Question 7


```

# Q8. [Optional]: Why might this data be missing?
They may have been vaccinated originally in a different county, so not all their records are available in San Diego 


Working with dates
```{r}
library(lubridate)

```

# Q9. How many days have passed since the last update of the dataset? 

```{r}
 today() - ymd(vax$as_of_date[nrow(vax)])
```

# Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
nrow(table(vax$as_of_date))
```


Working with zip codes 
```{r}
library(zipcodeR)
library(dplyr)

```
Using dplyr to look at San Diego vaccinations:

```{r}
sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Filtering for areas with population of over 10000:
```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

# Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```

# Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 
```{r}
which.max(sd$age12_plus_population)
sd$zip_code_tabulation_area[53]
```

# Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-11-15”?
Filtering with dplyr:
```{r}
sd.subset <- filter(vax, county == "San Diego" &
                as_of_date == "2022-11-15" )

mean.default(sd.subset$percent_of_population_fully_vaccinated, na.rm = T)*100
```

# Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-11-15”?

```{r}
hist.data <- sd.subset$percent_of_population_fully_vaccinated
hist(hist.data, main = "Histogram of Vaccination Rates Across San Diego County",
     xlab = "Percent of Population Fully Vaccinated on 2022-11-15")
```

Narrowing in on only La Jolla: 

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
# head(ucsd)
```

# Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:
```{r}
library(ggplot2)
plot.a <-  ggplot(ucsd) +
  aes(x = as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) + 
  labs(x = "Date", y="Percent Vaccinated") + theme(axis.text.x = element_blank())
plot.a
```

# # Subset to all CA areas with a population as large as 92037
```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                    as_of_date == "2022-11-15")
               
# head(vax.36)
```

# Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?
```{r}
mean.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
mean.36
```

```{r}
plot.a +
  geom_hline(yintercept = mean.36, col = "red", linetype =2)
```

# Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-11-15”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

# Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) + 
  aes(percent_of_population_fully_vaccinated) + 
  geom_histogram() + xlab("Percent Vaccinated") + ylab("Count") + ylim(0,60)
```

# Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
```{r}
vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date == "2022-11-15") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)


```
92040: below
92109: below

# Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
# head(vax.36.all)
mean.36.all <- mean.default(vax.36.all$percent_of_population_fully_vaccinated, na.rm = T)
mean.36.all*100
```

```{r}
ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(0,1) +
  labs(x= "Date", y= "Percent Vaccinated",
       title="Vaccination Rates Across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = mean.36.all, linetype = 2)
```

# Q21. How do you feel about traveling for Thanksgiving Break and meeting for in-person class afterwards?
I didn't travel for Thanksgiving, but I will test before coming to class. I really prefer in-person class. 





